---
title: "670 Project - Happiness in the Time of Covid"
output: html_notebook
---
We found some interesting datasets on Kaggle - namely the World Happiness Reports from2015-2019, and 2021. They contain information like how GDP and life expectancy contributeto a "happiness rating" (based on the Gallup World Poll). We also have some COVID-19 datafor all countries which includes cases per day, deaths and cumulative counts over 2020.

The three Kaggle datasets we used are at the following URLS:

* https://www.kaggle.com/tunguz/data-on-covid19-coronavirus?select=owid-covid-data.csv
* https://www.kaggle.com/ajaypalsinghlo/world-happiness-report-2021
* https://www.kaggle.com/unsdsn/world-happiness

Load the libraries we need for the project:
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(gtools)
```

Load the datasets:

```{r}
happiness_2015 <- read.csv("data/2015.csv",header=T,na.strings="?")
happiness_2016 <- read.csv("data/2016.csv",header=T,na.strings="?")
happiness_2017 <- read.csv("data/2017.csv",header=T,na.strings="?")
happiness_2018 <- read.csv("data/2018.csv",header=T,na.strings="?")
happiness_2019 <- read.csv("data/2019.csv",header=T,na.strings="?")
happiness_2021 <- read.csv("data/world-happiness-report-2021.csv",header=T,na.strings="?")

covid2020_filtered <- read.csv("data/covid2020-filtered.csv", header = T, na.strings = "?")
```

# 1. Data Preprocessing
Manually, once we loaded in the data we noticed many column name mismatches and had to rename quite a few (for example, "Country" vs "Country or Region" or "Score" vs. "Happiness Score"). We started renaming the columns programatically but it was very tedious code to write and look at so we took it out of the notebook, we show how we changed the country column name just as an example.

##### Rename "Country.or.region" to "Country" in 2018/2019 so we can join with the rest.
```{r}
colnames(happiness_2018)[2] <- "Country"
names(happiness_2018)
colnames(happiness_2019)[2] <- "Country"
names(happiness_2019)

happiness_2015 = subset(happiness_2015, select=-c(Region,Standard.Error,Dystopia.Residual))
happiness_2016 = subset(happiness_2016,select=-c(Region,Lower.Confidence.Interval,Upper.Confidence.Interval,Dystopia.Residual))
happiness_2017 = subset(happiness_2017, select=-c(Whisker.high,Whisker.low,Dystopia.Residual))
```
##### Add a "Year" Column to each to keep track before merging.
```{r}
happiness_2015$Year=2015
happiness_2016$Year=2016
happiness_2017$Year=2017
happiness_2018$Year=2018
happiness_2019$Year=2019
```

##### Combine 2015-2019 happiness data into one dataframe
```{r}
happiness = smartbind(happiness_2015, happiness_2016, happiness_2017, happiness_2018, happiness_2019)
dim(happiness)
names(happiness)
```
#### Dataset Explanation:
* There are six factors – economic production(GDP per capita), social support, life expectancy(Healthy life expectancy), freedom, absence of corruption(Perceptions of corruption), and generosity – contributing to the calculation of the happiness score of each country.

* We filtered out the total COVID-19 cases and total COVID-19 deaths by December 31st, 2020 of each country.

# 2. Models

### Models and Analysis on Previous Years' Data
#### Q: What factors seem to contribute to a "happier" country?
##### Check linear statistical significance of all predictors in response to happiness score.
```{r}
lm.score.1 = lm(Happiness.Score~ GDP.per.capita + Social.support + Healthy.life.expectancy + Freedom + Generosity + Perceptions.of.corruption, data = happiness)
summary(lm.score.1)

# Plot predictions and residuals
plot(predict(lm.score.1), residuals(lm.score.1))
```
##### Check linear statistical significance of all predictors in response to happiness rank. 

```{r}
lm.rank.1 = lm(Happiness.Rank~ GDP.per.capita + Social.support + Healthy.life.expectancy + Freedom + Generosity + Perceptions.of.corruption, data = happiness)
summary(lm.rank.1)

# Plot predictions and residuals
plot(predict(lm.rank.1), residuals(lm.rank.1))
```
##### Fit a linear model with significant predictors in response to happiness score. 

```{r}

lm.score.2 = lm(Happiness.Score~ GDP.per.capita + Healthy.life.expectancy + Freedom, data = happiness)
summary(lm.score.2)

# Plot predictions and residuals
plot(predict(lm.score.2), residuals(lm.score.2))
```

##### Fit a linear model with significant predictors in response to happiness rank.

```{r}
 
lm.rank.2 = lm(Happiness.Rank~ GDP.per.capita + Healthy.life.expectancy + Freedom, data = happiness)
summary(lm.rank.2)

# Plot predictions and residuals
plot(predict(lm.rank.2), residuals(lm.rank.2))
```
#### Results of linear models: 
* GDP per capita, Healthy life expectancy, Social support and Freedom seem to contribute the most to a "happier" country. 
* Generosity and Perceptions of corruption are the least significant predictors.
* However, Social support has been removed as not all previous data include such information.
* The plot of the rank models are the opposite of the score models.
* By removing generosity and perceptions of corruption, we get similar R-squared values but a higher F-statistics.

##### Fit a linear model with GDP.per.capita in response to happiness score using K-fold cross validation.
```{r}
library(boot)
set.seed(1)
cv.error.10 = rep(0:10)
glm.score = glm(Happiness.Score~GDP.per.capita, data = happiness)
cv.error.10 =cv.glm(happiness, glm.score, K=10)
cv.error.10$delta
```


##### Fit a linear model with Healthy life expectancy in response to happiness score using K-fold cross validation.
```{r}

glm.score.life = glm(Happiness.Score~Healthy.life.expectancy, data = happiness)
cv.error.10.life =cv.glm(happiness, glm.score.life, K=10)
cv.error.10.life$delta
```


##### Fit a linear model with Freedom in response to happiness score using K-fold cross validation.
```{r}

glm.score.free = glm(Happiness.Score~Freedom, data = happiness)
cv.error.10.free =cv.glm(happiness, glm.score.free, K=10)
cv.error.10.free$delta
```

#### Fun facts about the 1-1 relationships: 
* GDP per capita seems to have a clearer correlation with the happiness score, then healthy life expectancy, then freedom. 



##### Fit a linear model with multiple predictors in response to happiness score using K-fold cross validation.
```{r}
set.seed(8)

glm.score.multi = glm(Happiness.Score ~ GDP.per.capita +  Healthy.life.expectancy + Freedom, data = happiness)
cv.error.5.multi =cv.glm(happiness, glm.score.multi, K=5)
cv.error.5.multi$delta
```

#### The cross validation error improved!


### Models and Predictions on Data of Year 2021
#### Q: Does the impact of COVID-19 seem to have an effect on the happiness rating?

##### Merge happiness 2021 table with covid 2020 table using inner join.
```{r}

happy.covid <- happiness_2021 %>% inner_join(covid2020_filtered, by = c("Country.name" = "location"))

# statistics
summary(happy.covid[, c("Ladder.score", "Logged.GDP.per.capita", "Healthy.life.expectancy", "Freedom.to.make.life.choices", "total_cases", "total_deaths")])
```


##### Fit a linear model with happiness 2021 data only featuring significant predictors in response to happiness score.
```{r}
attach(happy.covid)

lm.score.wo.covid = lm(Ladder.score ~ Logged.GDP.per.capita + Social.support + Healthy.life.expectancy+ Freedom.to.make.life.choices + Generosity + Perceptions.of.corruption, data=happy.covid)
summary(lm.score.wo.covid)
```

##### Fit a linear model with both happiness and covid data with significant predictors in response to happiness score.
```{r}
lm.score.w.covid = lm(Ladder.score ~ Logged.GDP.per.capita + Social.support + Healthy.life.expectancy+ Freedom.to.make.life.choices + Generosity + Perceptions.of.corruption + total_cases + total_deaths, data=happy.covid)
summary(lm.score.w.covid)
```

#### Results from the above two models:
* Total covid cases and total deaths don't seem to affect the happiniess (Ladder) score much.

##### Predict happiness scores of 2021 using the models trained on previous years' data.
```{r}
# prepare test data
data.2021 <- data.frame(GDP.per.capita=Logged.GDP.per.capita,Healthy.life.expectancy = Healthy.life.expectancy, Freedom=Freedom.to.make.life.choices)
lm.pred.score = predict(lm.score.2, data.2021) # without cross validation
glm.pred.score = predict(glm.score.multi, data.2021) # with cross validation

# append pred score to the dataframe
happy.covid$lm.pred.score = lm.pred.score
happy.covid$glm.pred.score = glm.pred.score

# pred statistics
summary(happy.covid[, c('lm.pred.score', 'glm.pred.score')])
```

##### Predicted rank per linear model without cv
```{r}
lm.pred.country.score <- happy.covid[, c("Country.name", "lm.pred.score")]
lm.pred.rank = lm.pred.country.score[order(-lm.pred.score),]
lm.pred.rank
```

##### Predicted rank per glm model with cv
```{r}
glm.pred.country.score <- happy.covid[, c("Country.name", "glm.pred.score")]
glm.pred.rank = glm.pred.country.score[order(-glm.pred.score),]
glm.pred.rank

```

### Plot Predicted vs. Actual
```{r}
actual <- subset(happy.covid, select=c(Country.name, Ladder.score ))
colnames(actual)[2] <- "Score"
actual <- actual[order(-actual$Score),]
actual$Rank <- 1:nrow(actual)
actual <- actual[1:15, ]

lm_predicted <- subset(happy.covid, select=c(Country.name, lm.pred.score))
lm_predicted <- lm_predicted[order(-lm_predicted$lm.pred.score),]
lm_predicted$lm_Rank <- 1:nrow(lm_predicted)

glm_predicted <- subset(happy.covid, select=c(Country.name, glm.pred.score))
glm_predicted <- glm_predicted[order(-glm_predicted$glm.pred.score),]
glm_predicted$glm_Rank <- 1:nrow(glm_predicted)

results <- actual %>% inner_join(lm_predicted, by = "Country.name")
results <- results %>% inner_join(glm_predicted, by = "Country.name")

ggplot(results, aes(x=Country.name, colour=Country.name)) +
geom_point( aes(y=Rank), size=4) +
geom_point( aes(y=lm_Rank), size=4, shape=3) +
geom_point( aes(y=glm_Rank), size=4, shape=4) +
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
### Other Plots and Analysis

#### Q: How did the total deaths caused by COVID-19 correlate with the GDP of a country?
```{r}
p <- ggplot(covid2020_filtered, aes(gdp_per_capita, total_deaths)) 
p + geom_point()
# p + geom_point() + scale_x_continuous(trans='log2') +
#   scale_y_continuous(trans='log2')
p + geom_bin2d() + scale_fill_gradient(low="green", high="red")
```

#### Q: How did COVID-19 affect happiness ratings in developed countries compared to developing countries?
```{r}
cp <- ggplot(happy.covid, aes(total_deaths, Ladder.score)) 
cp + geom_point()
```

##### Happiness score of Canada, United States, and Mexico from 2015 to 2019.
```{r}
happiness_na <- subset(happiness, Country == "Canada" | Country == "United States" | Country =="Mexico",
select=c(Country, Year, Happiness.Score, Happiness.Rank))
ggplot(happiness_na, aes(Year, Happiness.Score, group=Country)) + geom_line(aes(color=Country))
```

##### Happiness rank of Canada, United States, and Mexico from 2015 to 2019.
```{r}
ggplot(happiness_na, aes(Year, Happiness.Rank, group=Country)) + geom_line(aes(color=Country))
```
##### Plot total COVID-19 cases and total deaths of each continent.
```{r}
dp <- ggplot(covid2020_filtered, aes(total_cases, total_deaths, colour = continent)) 
dp + geom_point()
```

# Discussion
* What factors seem to contribute to a "happier" country?
* Does the impact of COVID-19 seem to have an effect on the happiness rating?
* Does the impact of COVID-19 seem to have an effect on the amount certain factors contribute to happiness ratings?
* Do countries that were heavily impacted by COVID-19 have a lower happiness rating in 2021 compared to 2019?
* How did COVID-19 affect happiness ratings in developed countries compared to developing countries?
* Can we use additional data from previous years (2015-2019) to predict happiness ratings for 2021 and compare these with the actual?
